---
title: "Random forest training for 3'-end classification"
output:
  rmdformats::readthedown:
    toc_float: true
    highlight: kate
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align='center', dev = 'svg')
require(GenomicRanges)
require(GenomicFeatures)
require(dplyr) 
require(tidyr)
require(ggplot2)
require(Biostrings)
require(rtracklayer)
require(readr)
require(stringr)
require(caTools)
require(ranger)
require(patchwork)
require(memes)
require(gridExtra)
source("code/rf_helpers.R")
```

# Input data 

## Load reference annotation data and gencode data

```{r load reference annotation and genome, message=FALSE, warning=FALSE}
fastagenomegz <- "input/GRCh38_no_alt_analysis_set_GCA_000001405.15.fasta"
human_genome <- readDNAStringSet(fastagenomegz)
chromosome_names <- names(human_genome)
chromosome_sizes <- seqlengths(human_genome)
#clean chromosome names
for (i in seq_along(chromosome_names)) { 
  chromosome_names[i] <- unlist(strsplit(chromosome_names[i], " "))[1]
}
names(human_genome) <- chromosome_names

#load assembly
all.ends <- read.csv("input/n3_Neuron_THP1_iPSchro_all_internal_priming_exclude_Gencode.tsv", header=T, sep="\t")
colnames(all.ends)[1:6] <- c("chr","start","end","counts.1","counts.2","strand")

# Convert to GRanges object, keep relevant metadata
ends.custom.assembly.table1 <- GRanges(
  seqnames = all.ends$chr,
  ranges = IRanges(start = all.ends$end, end = all.ends$end),
  strand = all.ends$strand,
  metadata = DataFrame(all.ends[,c(5,7:10)])
)
rm(all.ends)

#remove TES entries exceeding chromosome boundaries +/- 150 nt
outofbound <- extend(ends.custom.assembly.table1, upstream = 150, downstream = 150)
outofbound_index_end <- as.vector(which(end(outofbound) > seqlengths(human_genome)[as.character(seqnames(outofbound))]))
outofbound_index_start <- which(start(outofbound) < 0)
outofbound_index <- c(outofbound_index_end,outofbound_index_start)
ends_trimmed <- ends.custom.assembly.table1[-outofbound_index]
rm(outofbound, outofbound_index, outofbound_index_end, outofbound_index_start, ends.custom.assembly.table1)
names(mcols(ends_trimmed)) <- gsub("metadata.", "", names(mcols(ends_trimmed)))

#load gencode
all.ends.gencode <- read.csv("input/n3_Gencode_internal_priming.tsv", header=T, sep="\t")
colnames(all.ends.gencode)[1:6] <- c("chr","start","end","counts.1","counts.2","strand")
# Convert to GRanges object, keep relevant metadata
ends_trimmed.gencode <- GRanges(
  seqnames = all.ends.gencode$chr,
  ranges = IRanges(start = all.ends.gencode$end, end = all.ends.gencode$end),
  strand = all.ends.gencode$strand,
  metadata = DataFrame(all.ends.gencode[,c(5,7:10)])
)
rm(all.ends.gencode)
#from gencode, no need to remove out of bound
names(mcols(ends_trimmed.gencode)) <- gsub("metadata.", "", names(mcols(ends_trimmed.gencode)))

```

### Clustering of assembly 3'ends

3'ends from assembly were clustered in windows of 5 nt, then the most distal 3'end was kept as the reference nucleotide of the cluster.  

```{r, make clusters from assembly, message=FALSE, warning=FALSE}
window = 0 #note tune parameter if want to reduce n. of 3' endby their proximity
# Reduce the ranges
reduced_gr <- reduce(ends_trimmed, min.gapwidth = window )
reduced_gr$cluster_id <- paste0("clusterID_", 
                                seqnames(reduced_gr), 
                                ":", 
                                start(reduced_gr),"-", end(reduced_gr),
                                "|",
                                strand(reduced_gr))
three_prime_ends <- ifelse(strand(reduced_gr) == "+", end(reduced_gr), start(reduced_gr))
three_prime_gr <- GRanges(
  seqnames = seqnames(reduced_gr),
  ranges = IRanges(
    start = three_prime_ends,
    width = 1
  ),
  strand = strand(reduced_gr)
)
three_prime_gr$cluster_id <- reduced_gr$cluster_id 

a <- plotNucleotides(three_prime_gr, human_genome) + ggtitle("All ONT-CAGE")+guides(fill=guide_legend(title="Nt"))+ylim(0,1)+theme_bw()+theme(legend.title=element_blank())

assembly_clusters <- three_prime_gr
rm(three_prime_gr)

#separately for gencode
# Reduce the ranges
reduced_gr.gencode <- reduce(ends_trimmed.gencode, min.gapwidth = window )
reduced_gr.gencode$cluster_id <- paste0("clusterID_", 
                                seqnames(reduced_gr.gencode), 
                                ":", 
                                start(reduced_gr.gencode),"-", end(reduced_gr.gencode),
                                "|",
                                strand(reduced_gr.gencode))
three_prime_ends.gencode <- ifelse(strand(reduced_gr.gencode) == "+", end(reduced_gr.gencode), start(reduced_gr.gencode))
three_prime_gr.gencode <- GRanges(
  seqnames = seqnames(reduced_gr.gencode),
  ranges = IRanges(
    start = three_prime_ends.gencode,
    width = 1
  ),
  strand = strand(reduced_gr.gencode)
)
three_prime_gr.gencode$cluster_id <- reduced_gr.gencode$cluster_id 

b <- plotNucleotides(three_prime_gr.gencode, human_genome)+ ggtitle("Gencode")+guides(fill=guide_legend(title="Nt"))+ylim(0,1)+theme_bw()+theme(legend.title=element_blank())

assembly_clusters.gencode <- three_prime_gr.gencode
rm(three_prime_gr.gencode)

```

## Import reference data

This is the data used for training the model, data comes from 1) FLAM-seq data + 3'-seq data in Human organoids (Alfonso-Gonzalez, C., et al., 2023) and FLAM-seq data from iPSC (Legnini, I., et al., 2019). 

### Load poly(A) databases 

### Prepare "neuronal" database (iPSC-derived cerebral organoids)

```{r import polyA database, message=FALSE, warning=FALSE}
path_neuronal_database <- "input/neuronal.gff"
neuronal_database <- rtracklayer::import.gff(path_neuronal_database)
end(neuronal_database) <- end(neuronal_database)+1 #fix 3'ends (0/1 based)
start(neuronal_database) <- start(neuronal_database)+1
```

### Prepare FLAM-seq from iPSCs

```{r, prepare poly(A) db, warning=FALSE, message=FALSE}
dirFlam <- "input"
flam_bams <- list.files(dirFlam, pattern = "*.bam", full.names = TRUE)
# Load bam files of genomic alignments 
flamSeqData <- unlist(GRangesList(lapply(flam_bams, function(x){
  GenomicAlignments::readGAlignments(x) %>% 
  GenomicRanges::GRanges(.) 
})))
# Invert strand 
strand(flamSeqData) <- ifelse(strand(flamSeqData) == '+', '-', '+') 
# Take last nucleotide of reads 
flam3pEnds<- trim_single_nucleotide(flamSeqData)
end(flam3pEnds)<-ifelse(strand(flam3pEnds)=="+", end(flam3pEnds)+1,end(flam3pEnds)-1) #fix 3'end (0/1 based) 
start(flam3pEnds)<-ifelse(strand(flam3pEnds)=="+", start(flam3pEnds)+1,start(flam3pEnds)-1)

# Put together both databases 
combined.database <- c(neuronal_database, flam3pEnds)
#cluster again 
combined.reduced <- reduce(combined.database,  min.gapwidth = 10) 
combined.reduced_ends <- ifelse(strand(combined.reduced) == "+", 
                                end(combined.reduced), 
                                start(combined.reduced))
combined.reduced.fromClusters <- GRanges(
    seqnames = seqnames(combined.reduced),
    ranges = IRanges(
        start = combined.reduced_ends,
        width = 1
    ),
    strand = strand(combined.reduced)
)
seqlevelsStyle(combined.reduced.fromClusters) <- "UCSC"
standard_chromosomes <- c(paste0("chr", 1:22), "chrX", "chrY")
# Keeping only ranges from standard chromosomes
combined.database <- combined.reduced.fromClusters[seqnames(combined.reduced.fromClusters) %in% standard_chromosomes]
```

# Nucleotide probabilities in reference database 
```{r, plot reference Database, warning=FALSE, message=FALSE}
c <- plotNucleotides(combined.database, BSgenome.Hsapiens.UCSC.hg38) + ggtitle("Reference (FLAM-seq + filtered 3p-seq)")+guides(fill=guide_legend(title="Nt"))+ylim(0,1)+theme_bw()+theme(legend.title=element_blank())

```

### Quality of reference datasets 

### Plot overlapping clusters with reference. 
```{r,message=FALSE, warning=FALSE}
assembly_clusters.inReference <- subsetByOverlaps(assembly_clusters, 
                                                  combined.database, 
                                                  maxgap =10) # note optimized window
d <- plotNucleotides(assembly_clusters.inReference, human_genome)+ggtitle("ONT-CAGE in Reference")+guides(fill=guide_legend(title="Nt"))+ylim(0,1)+theme_bw()+theme(legend.title=element_blank())

ntprobinput<-((a | b) / (c | d))
ntprobinput
ggsave(filename = "output/ntprob_input.pdf", plot = ntprobinput, device = "pdf", width = 8, height = 8, dpi = 300)

```

### Identify 3'-ends in reference 

```{r, annotate clusters, message=FALSE, warning=FALSE}
window = 10
assembly_clusters.inReference <- subsetByOverlaps(assembly_clusters, 
                                                  combined.database, 
                                                  window)
# add label for those in reference 
assembly_clusters$inRef <- ifelse(
  assembly_clusters$cluster_id %in% 
    assembly_clusters.inReference$cluster_id, 
  TRUE,
  FALSE)

#separately for gencode

assembly_clusters.inReference.gencode <- subsetByOverlaps(assembly_clusters.gencode, 
                                                  combined.database, 
                                                  window)
# add label for those in reference 
assembly_clusters.gencode$inRef <- ifelse(
  assembly_clusters.gencode$cluster_id %in% 
    assembly_clusters.inReference.gencode$cluster_id, 
  TRUE,
  FALSE)

```

### Retrieve  features to train the model or run it. 

```{r, extract sequence features, message=FALSE, warning=FALSE}
features_in_assembly.3pends <- extract_Seqfeatures(clusters3seq = assembly_clusters,
                                                   genomeSeq = human_genome, 
                                                   dwindow = 40,  #note optimized parameter
                                                   upwindow = 40)

features_in_assembly.3pends <- features_in_assembly.3pends %>%
  mutate(signalDistance = ifelse(is.na(signalDistance), 
                                 -99, 
                                 signalDistance))

#separately for gencode

features_in_assembly.3pends.gencode <- extract_Seqfeatures(clusters3seq = assembly_clusters.gencode,
                                                   genomeSeq = human_genome, 
                                                   dwindow = 40,  #note optimized parameter
                                                   upwindow = 40)

features_in_assembly.3pends.gencode <- features_in_assembly.3pends.gencode %>%
  mutate(signalDistance = ifelse(is.na(signalDistance), 
                                 -99, 
                                 signalDistance))

```


### Features (PAS) associated with 3'ends (in vs not in Reference) 

```{r,message=FALSE, warning=FALSE}

#roerder PASs

features_in_assembly.3pends$polyA.Signal <- factor(features_in_assembly.3pends$polyA.Signal, levels = c("AATAAA","ATTAAA","AATATA","TATAAA", "TTTAAA", "AATACA","AAGAAA", "CATAAA","AATGAA","GATAAA","ACTAAA","AATAGA","not detected"))

features_in_assembly.3pends.gencode$polyA.Signal <- factor(features_in_assembly.3pends.gencode$polyA.Signal, levels = c("AATAAA","ATTAAA","AATATA","TATAAA", "TTTAAA", "AATACA","AAGAAA", "CATAAA","AATGAA","GATAAA","ACTAAA","AATAGA","not detected"))

e <- ggplot(features_in_assembly.3pends) +
  geom_bar(aes(x=inRef, 
               fill=polyA.Signal),position="fill") + 
  theme_classic() + 
  ylab("Fraction of 3' ends") + 
  xlab("TES in Reference") +
  theme(text = element_text(size=20))+ 
  scale_fill_viridis_d()+
  ggtitle("ONT-CAGE")+
  guides(fill=guide_legend(title="PAS motif"))

f <- ggplot(features_in_assembly.3pends.gencode) +
  geom_bar(aes(x=inRef, 
               fill=polyA.Signal),position="fill") + 
  theme_classic() + 
  ylab("Fraction of 3' ends") + 
  xlab("TES in Reference") +
  theme(text = element_text(size=20))+ 
  scale_fill_viridis_d()+
  ggtitle("Gencode")+
  guides(fill=guide_legend(title="PAS motif"))

pasinput<-(e | f)
pasinput
ggsave(filename = "output/pas_input.pdf", plot = pasinput, device = "pdf", width = 12, height = 8, dpi = 300)

```

# Random forest model training (skip if pretrained model is available)

## Prepare training and test sets

Due to the significantly larger number of non-overlapping sets compared to the reference, we utilized a smaller sampling proportion of 3' ends not found in the reference for training, in order to balance the dataset, sampling a fraction 0.8 of TRUE positives and 0.2 of the set of 3'ends not found in reference. 

```{r, prepare training and test datasets, message=FALSE, warning=FALSE}
# make training sets 
positive <- features_in_assembly.3pends %>% filter(inRef==TRUE)
negative <- features_in_assembly.3pends %>% filter(!inRef==TRUE)
# sampling
split_positive <- sample.split(positive, SplitRatio = 0.8)
split_negative <- sample.split(negative, SplitRatio = 0.2)
#### trin set 
# subset
positive_train <- subset( positive , split_positive == "TRUE")
negative_train <- subset( negative, split_negative == "TRUE")
# training
train <- rbind(positive_train, 
               negative_train)
train$inRef <- as.factor(train$inRef)
rownames(train) <- make.unique( train$cluster_id) 
train <-
  train %>% dplyr::select(
    -c(
      cluster_id
    )
  ) %>%
  mutate(signalDistance = ifelse(is.na(signalDistance), 
                                 -99, 
                                 signalDistance))

rownames(features_in_assembly.3pends) <- make.unique( features_in_assembly.3pends$cluster_id ) 
# Testing data 
test_data <- features_in_assembly.3pends %>% 
  filter(!cluster_id %in% rownames(train)) %>% 
  dplyr::select(
    -c(
      cluster_id
    )
  ) %>%
  mutate(signalDistance = ifelse(is.na(signalDistance), 
                                 -99, 
                                 signalDistance))

```

## Traning parameters (skip if pretrained model is available)

In our analysis, we utilized the `ranger` function to develop a Random Forest classifier, named `classifier_mtry10`, trained on the `train` dataset with `inRef` as the classification variable and all other variables as predictors. The model construction involved generating 1000 trees (`num.trees = 1000`), each considering a random subset of 10 predictors at every split (`mtry = 10`), and assessing feature importance based on impurity reduction (`importance = "impurity"`). We also addressed potential class imbalance and refined our probability threshold for enhanced accuracy and sensitivity by applying weights (`weights <- c(FALSE = 0.03, TRUE = 1)`) via the `class.weights` parameter. Additionally, the `probability = TRUE` parameter was activated for obtaining detailed class probabilities. This approach, especially given our clear a priori readouts (polyA signals and nucleotide distributions), enabled us to determine an optimal probability cutoff, effectively balancing accuracy and sensitivity in the model. The `seed` parameter was set to `NULL` to allow for random seeding, contributing to the model’s robustness. 

```{r, training parameters}
modelDir <- "output/rf_classifier.RDS"
weights <- c(`FALSE` = 0.03, `TRUE` = 1)
rf_classifier = ranger(inRef   ~ ., 
                       data=train, 
                       num.trees = 500, 
                       mtry=10, 
                       importance = "impurity",
                       seed = 42,
                       probability = TRUE,
                       class.weights = weights,
                       max.depth = 20) 
saveRDS(rf_classifier, modelDir)
```

# Features of importance (skip if pretrained model is available)

```{r}
as.data.frame( rf_classifier$variable.importance ) %>% 
  dplyr::mutate(feature = rownames(.)) %>%
  dplyr::rename(importance = 1) %>% 
  dplyr::mutate(importance = as.numeric(importance)) %>%
  filter(importance>0) %>%
  arrange(desc(importance)) %>% 
  ggplot(data=., aes(x=reorder(feature, importance) , 
                     y=importance)) +
  geom_bar(stat="identity", fill="dodgerblue") + 
  coord_flip() + 
  theme_classic() +
  theme(legend.position = "none") +
  xlab("Features") + 
  ylab("Importance")

```

### Accuracy and sensitivity on test data: recovered 3'ends per cut-off (skip if pretrained model is available)

```{r, predict from test set}
# Load pre-trained model with above parameters 
t1 <- path.rf_model <- modelDir
rf_model <- readRDS(path.rf_model)
model.predictions <- predict(rf_model, data = test_data)

# set TRUE classification probabilities 
cut_offs <- c(0.001,0.01,0.05,0.1,0.25,0.5,0.75,0.95)
names(cut_offs) <- paste0( ">", cut_offs) 
# Plot recovered 3'ends per cut-off 
t1<-lapply(cut_offs, function(x) {
  model.predictions$predictions %>% 
    as.data.frame(.) %>% 
    filter(`TRUE` > x) %>% 
    nrow(.)
}) %>% 
  unlist(.) %>% 
  as.data.frame(.) %>%
  dplyr::rename(num_of_3_ends = 1) %>%
  mutate(probability = gsub("prob", "", rownames(.)))  %>%
  ggplot(., aes(fill = probability, y = num_of_3_ends, x = probability)) +
  geom_bar(position = "dodge", stat = "identity") + theme_classic() + 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))+
  scale_fill_brewer(palette="RdBu")

```

### Recovered poly(A) signals per cut-off (skip if pretrained model is available)

```{r, recovered polyA per cut-off}
test_data$prediction_prob <- model.predictions$predictions[,2]
t2 <- lapply(cut_offs, function(x){
  test_data %>% 
    mutate(predicted_true = ifelse(prediction_prob>x, TRUE,FALSE)) %>% 
    group_by(polyAsignalClass, predicted_true) %>% 
    tally() %>% 
    group_by(predicted_true) %>%
    mutate(percentage = n / sum(n)) %>% 
    as.data.frame(.)}) %>% 
  do.call(rbind,.) %>% 
  mutate(probability = paste0("0." , 
                              str_split_fixed(rownames(.), 
                                              "\\.", 
                                              n=3)[,2])) %>%
  ggplot(., aes(fill=polyAsignalClass, 
                y=percentage, 
                x=predicted_true)) + 
  geom_bar(position="stack", 
           stat="identity") +
  theme_classic() + 
  facet_grid(cols=vars(probability)) + 
  theme(axis.text.x = element_text(angle = 45, 
                                   vjust = 1, 
                                   hjust=1))+
  scale_fill_viridis_d()

```

### Sensitivity (skip if pretrained model is available)

```{r, sensitivity analysis}
t3 <- lapply(cut_offs, function(x){
  test_data %>% 
    mutate(predicted_true = ifelse(prediction_prob>x, TRUE,FALSE)) %>% 
    group_by(inRef, predicted_true) %>% 
    tally() %>% 
    group_by(predicted_true) %>%
    mutate(percentage = n / sum(n)) %>% 
    as.data.frame(.)}) %>% 
  do.call(rbind,.) %>% 
  mutate(probability = paste0("0." , str_split_fixed(rownames(.), "\\.", n=3)[,2])) %>%
  ggplot(., aes(fill=inRef, y=percentage, x=predicted_true)) + 
  geom_bar(position="stack", stat="identity") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  facet_grid(cols=vars(probability))+
  scale_fill_manual(values=c("dodgerblue","firebrick"))

qctest<-(t1 | t2 | t3)
qctest
ggsave(filename = "output/qc_testset.pdf", plot = qctest, device = "pdf", width = 16, height = 4, dpi = 300)

```

### Nucleotide compositions per cut-off (skip if pretrained model is available)

```{r, calculating nucleotide composition}
nt.probs <- lapply(cut_offs, function(x){
  true_probabilities <- model.predictions$predictions %>% 
    as.data.frame(.) %>% dplyr::pull(2)
  test_data$true_probabilities <- true_probabilities 
  class.true.clusters <- test_data %>% 
    mutate(cluster_id = rownames(.)) %>% 
    filter(true_probabilities>x) %>% 
    dplyr::pull(cluster_id)
  class.false.clusters <- test_data %>% 
    mutate(cluster_id = rownames(.)) %>% 
    filter(!true_probabilities>x) %>% dplyr::pull(cluster_id)
  res <- list()
  message("Found in cut off:",  x  , " ","found: ", length(assembly_clusters[assembly_clusters$cluster_id %in% class.true.clusters]), " ","positive 3'-ends")
  res$true.3p_ends <- plotNucleotides(  assembly_clusters[assembly_clusters$cluster_id %in% class.true.clusters] , human_genome) 
  res$false.3p_ends <- plotNucleotides(  assembly_clusters[assembly_clusters$cluster_id %in% class.false.clusters] , human_genome)
  return(res)
})

ntp1 <-  nt.probs$`>0.001`$true.3p_ends +ggtitle("Prob > 0.001")+ylim(0,1)+theme_bw()
ntp2 <-  nt.probs$`>0.01`$true.3p_ends +ggtitle("Prob > 0.01")+ylim(0,1)+theme_bw()
ntp3 <-  nt.probs$`>0.1`$true.3p_ends +ggtitle("Prob > 0.1")+ylim(0,1)+theme_bw()
ntp4 <-  nt.probs$`>0.25`$true.3p_ends +ggtitle("Prob > 0.25")+ylim(0,1)+theme_bw()

nttest <- ((ntp1 | ntp2) / (ntp3 | ntp4))
nttest
ggsave(filename = "output/ntprob_testset.pdf", plot = nttest, device = "pdf", width = 8, height = 8, dpi = 300)

```

Define thresholds based on fdr 5%

```{r, annotate clusters to assembly}

#find threshold of probability so that 95% of test inRef and PAS detected are kept
inRef.test_data <- test_data[test_data$inRef == TRUE & test_data$polyA.Signal != "not detected", ]
inRef.test_data.sorted <- inRef.test_data[order(inRef.test_data$prediction_prob, decreasing = TRUE), ]
true_index <- ceiling(0.95 * nrow(inRef.test_data))
true_threshold <- inRef.test_data.sorted$prediction_prob[true_index] #0.3465253
true_threshold
nrow(test_data[test_data$prediction_prob>true_threshold,])/nrow(test_data) #-> ca. 14.6% are significant poly(A)

#find threshold of probability so that 95% of non inRef no PAS are kept
noRef.test_data <- test_data[test_data$inRef == FALSE, ]
noRef.test_data.sorted <- noRef.test_data[order(noRef.test_data$prediction_prob, decreasing = TRUE), ]
false_index <- ceiling(0.95 * nrow(noRef.test_data))
false_threshold <- noRef.test_data.sorted$prediction_prob[false_index] #0.0001004803
false_threshold
nrow(test_data[test_data$prediction_prob<false_threshold,])/nrow(test_data) #ca. 4.8% are significant non-poly(A)


```
# Output results 

Output results with a column of probabilities 

## Annotate clusters to assembly 3'ends (start here if pretrained model is available)

```{r, annotate clusters to assembly}
#load pretrained model
modelDir <- "output/rf_classifier.RDS"
path.rf_model <- modelDir
rf_model <- readRDS(path.rf_model)

#run classification
rownames(features_in_assembly.3pends) <-
  make.unique(features_in_assembly.3pends$cluster_id)
model.predictions <-
  predict(rf_model, data = features_in_assembly.3pends)
# overlapping segments to clusters
features_in_assembly.3pends$probability_TRUE <-
  model.predictions$predictions[, 2]
assembly_clusters.annotated <- reduced_gr %>% 
  as.data.frame(.) %>%
  left_join(
    .,
    features_in_assembly.3pends %>%
      dplyr::select(polyAsignalClass, 
                    polyA.Signal, 
                    cluster_id, 
                    probability_TRUE,
                    inRef), by = "cluster_id") %>% 
  makeGRangesFromDataFrame(., keep.extra.columns = TRUE)
# findOverlaps 
hits_to_clusters <- findOverlaps( ends_trimmed, 
                                  assembly_clusters.annotated)
annotated_clusters <- ends_trimmed[queryHits(hits_to_clusters)]
mcols(annotated_clusters) <- mcols(assembly_clusters.annotated[subjectHits(hits_to_clusters)])

#separately for gencode
rownames(features_in_assembly.3pends.gencode) <-
  make.unique(features_in_assembly.3pends.gencode$cluster_id)
model.predictions.gencode <-
  predict(rf_model, data = features_in_assembly.3pends.gencode)
# overlapping segments to clusters
features_in_assembly.3pends.gencode$probability_TRUE <-
  model.predictions.gencode$predictions[, 2]
assembly_clusters.annotated.gencode <- reduced_gr.gencode %>% 
  as.data.frame(.) %>%
  left_join(
    .,
    features_in_assembly.3pends.gencode %>%
      dplyr::select(polyAsignalClass, 
                    polyA.Signal, 
                    cluster_id, 
                    probability_TRUE,
                    inRef), by = "cluster_id") %>% 
  makeGRangesFromDataFrame(., keep.extra.columns = TRUE)
# findOverlaps 
hits_to_clusters.gencode <- findOverlaps( ends_trimmed.gencode, 
                                  assembly_clusters.annotated.gencode)
annotated_clusters.gencode <- ends_trimmed.gencode[queryHits(hits_to_clusters.gencode)]
mcols(annotated_clusters.gencode) <- mcols(assembly_clusters.annotated.gencode[subjectHits(hits_to_clusters.gencode)])

#add frd thresholds-based class
annotated_clusters$class <- ifelse(annotated_clusters$probability_TRUE >= true_threshold, "significant poly(A)", "unknown / non-significant")
annotated_clusters[annotated_clusters$probability_TRUE <= false_threshold,]$class <- "significant non-poly(A)"
annotated_clusters.gencode$class <- ifelse(annotated_clusters.gencode$probability_TRUE >= true_threshold, "significant poly(A)", "unknown / non-significant")
annotated_clusters.gencode[annotated_clusters.gencode$probability_TRUE <= false_threshold,]$class <- "significant non-poly(A)"

```

# fraction of annotated positive based on ref per cut-off

```{r}
probability_cut_off <- c(0.001,0.01,0.05,0.1,0.25,0.5,0.75,0.95)
names(probability_cut_off) <- paste0(">", probability_cut_off)
t4 <- lapply(probability_cut_off, function(x){
  a1 <- annotated_clusters %>% as.data.frame(.) %>%
    filter(inRef==TRUE) %>% 
    mutate(isFound_inRef_and_Forest = ifelse(inRef==TRUE & probability_TRUE>x, TRUE, FALSE)) %>%
    group_by(isFound_inRef_and_Forest) %>% 
    tally() %>%
    mutate(percentage = n/sum(n)) %>%
    as.data.frame(.) 
  a1$probability <- x
  return(a1)
}) %>% do.call(rbind,.) %>%
  ggplot(., aes(fill=isFound_inRef_and_Forest, y=percentage, x=as.factor(probability))) + 
    geom_bar(position="stack", stat="identity") + 
  theme_classic()+ggtitle("ONT-CAGE in Ref.")+
  xlab("Cutoff")+
  scale_fill_manual(values=c("dodgerblue","firebrick"))+theme(legend.title=element_blank())
#separately for gencode
t5 <- lapply(probability_cut_off, function(x){
  a1 <- annotated_clusters.gencode %>% as.data.frame(.) %>%
    filter(inRef==TRUE) %>% 
    mutate(isFound_inRef_and_Forest = ifelse(inRef==TRUE & probability_TRUE>x, TRUE, FALSE)) %>%
    group_by(isFound_inRef_and_Forest) %>% 
    tally() %>%
    mutate(percentage = n/sum(n)) %>%
    as.data.frame(.) 
  a1$probability <- x
  return(a1)
}) %>% do.call(rbind,.) %>%
  ggplot(., aes(fill=isFound_inRef_and_Forest, y=percentage, x=as.factor(probability))) + 
    geom_bar(position="stack", stat="identity") + 
  theme_classic()+ggtitle("Gencode in Ref.")+
  xlab("Cutoff")+
  scale_fill_manual(values=c("dodgerblue","firebrick"))+theme(legend.title=element_blank())

t6 <- lapply(probability_cut_off, function(x){
  a1 <- annotated_clusters %>% as.data.frame(.) %>%
    filter(inRef==FALSE) %>% 
    mutate(isFound_in_Forest = ifelse(inRef==FALSE & probability_TRUE>x, TRUE, FALSE)) %>%
    group_by(isFound_in_Forest) %>% 
    tally() %>%
    mutate(percentage = n/sum(n)) %>%
    as.data.frame(.) 
  a1$probability <- x
  return(a1)
}) %>% do.call(rbind,.) %>%
  ggplot(., aes(fill=isFound_in_Forest, y=percentage, x=as.factor(probability))) + 
    geom_bar(position="stack", stat="identity") + 
  theme_classic()+ggtitle("ONT-CAGE all")+
  xlab("Cutoff")+
  scale_fill_manual(values=c("dodgerblue","firebrick"))+theme(legend.title=element_blank())

#separately for gencode
t7 <- lapply(probability_cut_off, function(x){
  a1 <- annotated_clusters.gencode %>% as.data.frame(.) %>%
       filter(inRef==FALSE) %>% 
    mutate(isFound_in_Forest = ifelse(inRef==FALSE & probability_TRUE>x, TRUE, FALSE)) %>%
    group_by(isFound_in_Forest) %>% 
    tally() %>%
    mutate(percentage = n/sum(n)) %>%
    as.data.frame(.) 
  a1$probability <- x
  return(a1)
}) %>% do.call(rbind,.) %>%
  ggplot(., aes(fill=isFound_in_Forest, y=percentage, x=as.factor(probability))) + 
    geom_bar(position="stack", stat="identity") + 
  theme_classic()+ggtitle("Gencode all")+
  xlab("Cutoff")+
  scale_fill_manual(values=c("dodgerblue","firebrick"))+theme(legend.title=element_blank())

```

### Recovered poly(A) signals per cut-off (skip if pretrained model is available)

```{r, recovered polyA per cut-off}
t8 <- lapply(cut_offs, function(x){
  as.data.frame(annotated_clusters) %>% 
    mutate(predicted_true = ifelse(probability_TRUE>x, TRUE,FALSE)) %>% 
    group_by(polyAsignalClass, predicted_true) %>% 
    tally() %>% 
    group_by(predicted_true) %>%
    mutate(percentage = n / sum(n)) %>% 
    as.data.frame(.)}) %>% 
  do.call(rbind,.) %>% 
  mutate(probability = paste0("0." , 
                              str_split_fixed(rownames(.), 
                                              "\\.", 
                                              n=3)[,2])) %>%
  ggplot(., aes(fill=polyAsignalClass, 
                y=percentage, 
                x=predicted_true)) + 
  geom_bar(position="stack", 
           stat="identity") +
  theme_classic() + 
  facet_grid(cols=vars(probability)) + 
  theme(axis.text.x = element_text(angle = 45, 
                                   vjust = 1, 
                                   hjust=1))+
  scale_fill_viridis_d()+ggtitle("ONT-CAGE")

t9 <- lapply(cut_offs, function(x){
  as.data.frame(annotated_clusters.gencode) %>% 
    mutate(predicted_true = ifelse(probability_TRUE>x, TRUE,FALSE)) %>% 
    group_by(polyAsignalClass, predicted_true) %>% 
    tally() %>% 
    group_by(predicted_true) %>%
    mutate(percentage = n / sum(n)) %>% 
    as.data.frame(.)}) %>% 
  do.call(rbind,.) %>% 
  mutate(probability = paste0("0." , 
                              str_split_fixed(rownames(.), 
                                              "\\.", 
                                              n=3)[,2])) %>%
  ggplot(., aes(fill=polyAsignalClass, 
                y=percentage, 
                x=predicted_true)) + 
  geom_bar(position="stack", 
           stat="identity") +
  theme_classic() + 
  facet_grid(cols=vars(probability)) + 
  theme(axis.text.x = element_text(angle = 45, 
                                   vjust = 1, 
                                   hjust=1))+
  scale_fill_viridis_d()+ggtitle("Gencode")

qc_cutoff<-((t4|t5)/(t6|t7)/(t8|t9))
qc_cutoff
ggsave(filename = "output/qc_bycutoff.pdf", plot = qc_cutoff, device = "pdf", width = 12, height = 12, dpi = 300)

```

### Nucleotide compositions per cut-off (skip if pretrained model is available)

```{r, calculating nucleotide composition}
nt.probs <- lapply(cut_offs, function(x){
  res <- plotNucleotides(annotated_clusters[annotated_clusters$probability_TRUE > x] , human_genome) 
  return(res)
})
ntp5 <-  nt.probs$`>0.001` +ggtitle("ONT-CAGE Prob > 0.001")+ylim(0,1)+theme_bw()
ntp6 <-  nt.probs$`>0.01` +ggtitle("ONT-CAGE Prob > 0.01")+ylim(0,1)+theme_bw()
ntp7 <-  nt.probs$`>0.1` +ggtitle("ONT-CAGE Prob > 0.1")+ylim(0,1)+theme_bw()
ntp8 <-  nt.probs$`>0.25` +ggtitle("ONT-CAGE Prob > 0.25")+ylim(0,1)+theme_bw()
nt.probs.gencode <- lapply(cut_offs, function(x){
  res <- plotNucleotides(annotated_clusters.gencode[annotated_clusters.gencode$probability_TRUE > x] , human_genome) 
  return(res)
})
ntp9 <-  nt.probs.gencode$`>0.001` +ggtitle("Gencode Prob > 0.001")+ylim(0,1)+theme_bw()
ntp10 <-  nt.probs.gencode$`>0.01` +ggtitle("Gencode Prob > 0.01")+ylim(0,1)+theme_bw()
ntp11 <-  nt.probs.gencode$`>0.1` +ggtitle("Gencode Prob > 0.1")+ylim(0,1)+theme_bw()
ntp12 <-  nt.probs.gencode$`>0.25` +ggtitle("Gencode Prob > 0.25")+ylim(0,1)+theme_bw()

nt_cutoff<-((ntp5 | ntp6 | ntp7 | ntp8)/(ntp9 | ntp10 | ntp11 | ntp12))
nt_cutoff
ggsave(filename = "output/ntprob_bycutoff.pdf", plot = nt_cutoff, device = "pdf", width = 16, height = 8, dpi = 300)

```

## Export files 
```{r}
rtracklayer::export.gff2(annotated_clusters, "output/annotated.3pclusters.gtf")
rtracklayer::export.gff2(annotated_clusters.gencode, "output/annotated.3pclusters.gencode.gtf")
```

# Session info 

```{r}
sessionInfo()
```
